Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-10-22 popl-rebuttal.md

- Decidability: The RG-TXN (Sec. 5.2, p. 20) rule does not lead to
  second-order quantification or quantified formulas with negation at
  prenex position. Sets are encoded as uninterpreted functions (Fig.
  10), and quantified set variables are simply replaced with new
  function symbols that occur free in the encoded formula.  Consider
  the ${\sf F} \Rightarrow G$ premise of RG-TXN. The quantified set
  variables $\delta$ and $\Delta$ are instantiated with fresh function
  variables $f_{\delta}$ and $f_{\Delta}$, respectively, resulting in
  a quantifier-free implication with free function symbols. The
  antecedent of this implication (${\sf F}$) is asserted as such, but
  its consequent ($G$) is negated before checking the satisfiability
  of the entire formula. This is a standard approach often employed to
  discharge verification conditions of the form $\forall a. \phi_1(a)
  \Rightarrow \phi_2(a)$.  In this process: (i). sets or functions are
  never quantified (thus no second-order quantification), and
  (ii). ${\sf F}$ is never negated (thus no inversion of quantifiers
  in ${\sf F}$). Other premises are similarly encoded.

  The reviewer is correct in observing that our encoding is not in the
  EPR fragment of FOL.  However, our encoding is still decidable as it
  belongs to a closely related variant of EPR, Godel-Kalmar-Schutte,
  that admits sentences of the form:
  $\exists^{*}\forall^{2}\exists^{*}$ (see: Borger et al, The
  Classical Decision Problem, 1st Edition, p.261, and Gurevich, On the
  Classical Decision Problem, p.7); formulas generated by our encoding
  clearly belong to this fragment.  We will clarify this issue in the
  final version.

- Local Context Independence: This assumption is used by the inference
  rule for `FOREACH`. The rule computes the transformer of a loop by
  lifting the transformer (${\sf F}$) of its body. In the process, it
  assumes that the effect of each loop iteration depends only on
  the loop variable $z$, but not on local ($\delta$) and global
  ($\Delta$) states (observe that ${\sf F}$ is applied to the same
  $\delta$ and $\Delta$ forall $z$). While independence on $\Delta$ is
  guaranteed by the stability condition (checked separately), the
  independence on the local state $\delta$ is simply assumed.

@@

Reviewer C
-------------

- Dealing with heap state: It is indeed true that we have ignored heap
  memory to focus exclusively on database state in our
  formalization. Swamy et al. (PLDI'13), however, have proposed the
  Dijkstra monad to reason about heap memory monadically.  As part of
  future work, we intend to explore the composition of the DB and
  Dijkstra monads to enable reasoning over both heap and database
  state.

Shepard
----------
- The primary concern raised by reviewer B is that the fragment of FOL
  in which the translation algorithm (Fig. 10, Sec.5) encodes set
  expressions is not EPR, hence its decidability is doubtful. In
  particular, the concern was that an existential quantifier follows a
  universal quantifier in the logical encoding of bind (`>>=`), and
  this pattern is not admitted by EPR. In the rebuttal phase we
  addressed the concern by pointing out that the fragment, while not
  EPR as claimed, is nonetheless decidable since it is subsumed by the
  GKS fragment, which is known to be decidable. However, as the
  reviewer rightly points out, there are no known results on the
  combination of GKS and Simple Linear Arithmetic (SLA) to support
  theorem 5.3, so SLA invariants, which are common in databases, might
  make the encoding undecidable. While this concern can be addressed
  by considering machine (finite-bit) arithmetic instead of SLA, we
  decided to take a long, hard look at the encoding algorithm and see
  if it can be simplified. In this we were successful as we were able
  to modify our bind encoding to eliminate the existential altogether. Our
  modifications are based on the observation that the semantics of
  bind, which were previously written using two implications one of
  which involves an existential, can be equivalently written using a
  single bi-implication that only involves universal quantifiers. A
  formal proof to this effect is given below, while an automatic
  Z3-assisted proof can be found [here](https://rise4fun.com/Z3/G1a).
  As a consequence of this change, the fragment of encoding is back to
  being EPR, hence the Theorems 5.2 and 5.3 continue to hold. We have
  also tightened our treatment of free variables in the encoding
  algorithm, and added an example to Sec. 5.2 to demonstrate how
  exactly the encoding works.

#### Bind Theorem and Proof ####

Theorem:
  Let S be a set containing elements of type T, and let R be a binary
  relation over T. Let S1 be a T set defined using the following two
  implications (note: (x,y: T) is a shorthand for (x:T, y:T)):

           ∀(x,y: T). x∈S ∧ R(x,y) ⇒ y∈S1			(Call it H0)
           ∀(y: T). y∈S1 ⇒ ∃(x:T). x∈S ∧ R(x,y)		(H1)

  Likewise, let S2 be a T set defined via a single bi-implication as
  following:

          ∀(x,y: T). x∈S ∧ R(x,y) ⇔ y∈S2			(H2)

  Then, S1 = S2. That is:

          ∀(y: T). y∈S1 ⇔ y∈S2        (G)

Proof:
  We will show G be proving the following two implications (goals)
  separately:

        ∀(y: T). y∈S1 ⇒ y∈S2					(G0)
        ∀(y: T). y∈S2 ⇒ y∈S1					(G1)

  Let us first note that H2 is a conjunction of the following two
  hypothesis:

          ∀(x,y: T). x∈S ∧ R(x,y) ⇒ y∈S2			(H3)
          ∀(x,y: T). y∈S2 ⇒ x∈S ∧ R(x,y)			(H4)

  Proof of G0:
  Intros on the goal G0 introduces a new y0:T, gives us the following
  hypothesis:

          y0∈S1							(H5)

  And reduces the goal to proving that y0∈S2.  Instantiating H1 with
  (y:=y0), and then applying it on H5 gives:

          ∃(x:T). x∈S ∧ R(x,y0)				(H6)

  Destructing the existential in H6 introduces a new x0:T, and gives the
  following:

          x0∈S ∧ R(x0,y0)				(H7)

  Now, instantiating H3 with (x:=x0)(y:=y0), and applying it on H7 gives
  y0∈S2, which is what needs to be proven.

  Proof of G1:
  Intros on the goal G1 introduces a new y0:T, gives us the following
  hypothesis:

          y0∈S2					(H8)

  And reduces the goal to proving that y0∈S1. At this point we prove a
  simple lemma that asserts the non-emptiness of type T:

          ∃(x:T)			(Lemma1)

  The proof for Lemma1 requires an evidence for a value of type T, which
  is readily available in the form of y0:T.  Destructing the existential
  in Lemma1 introduces a new x0:T into the context (note that x0 may or
  may not be equal to y0). Now instantiating H4 with (x:=x0)(y:=y0), and
  then applying it on H8 gives:

          x0∈S ∧ R(x0,y0)				(H9)

  Finally, instantiating H0 with (x:=x0)(y:=y0), and applying it on H9
  gives y0∈S1, which is what needs to be proven.
Qed.

### 2017-10-22 popl-rebuttal.md

  Now, instantiating H3 with (x:=x0)(y:=y0), and applying it on H7 gives
  y0∈S2, which is what needs to be proven.

  Proof of G1:
  Intros on the goal G1 introduces a new y0:T, gives us the following
  hypothesis:

          y0∈S2					(H8)

  And reduces the goal to proving that y0∈S1. Instantiating H4 with
  (x:=y0)(y:=y0), and then applying it on H8 gives:

          y0∈S ∧ R(y0,y0)				(H9)

  Finally, instantiating H0 with (x:=y0)(y:=y0), and applying it on H9
  gives y0∈S1, which is what needs to be proven.
Qed.

### 2017-10-22 popl-rebuttal.md

- Local Context Independence: This assumption is used by the inference
  rule for `FOREACH`. The rule computes the transformer of a loop by
  lifting the transformer (${\sf F}$) of its body. In the process, it
  assumes that the effect of each loop iteration depends only on
  the loop variable $z$, but not on local ($\delta$) and global
  ($\Delta$) states (observe that ${\sf F}$ is applied to the same
  $\delta$ and $\Delta$ forall $z$). While independence on $\Delta$ is
  gauranteed by the stability condition (checked separately), the
  independence on the local state $\delta$ is simply assumed.

Reviewer C
-------------

- Dealing with heap state: It is indeed true that we have ignored heap
  memory to focus exclusively on database state in our
  formalization. Swamy et al. (PLDI'13), however, have proposed the
  Dijkstra monad to reason about heap memory monadically.  As part of
  future work, we intend to explore the composition of the DB and
  Dijkstra monads to enable reasoning over both heap and database
  state.

Shepherd ----------

Dear Peter -

Please find attached a revised near-final version of our POPL
submission.  As per your guidance, while we have addressed the
comments from reviewers (A) and (C), we have taken special care
to consider the issues raised by reviewer (B).

The primary concern raised by the reviewer is that the fragment of FOL
in which the translation algorithm (Fig. 10, Sec.5) encodes set
expressions is not EPR, and hence raises doubts about its
decidability.  In particular, the concern was that an existential
quantifier follows a universal quantifier in the logical encoding of
bind (`>>=`), and this pattern is not admitted by EPR. In our
rebuttal, we addressed the concern by pointing out that the fragment,
while not EPR as claimed, is nonetheless decidable since it is
subsumed by GKS (Godel-Kalmar-Schutte), which is known to be
decidable. However, as the reviewer correctly points out, there are no
known results on the combination of GKS and Simple Linear Arithmetic
(SLA) to support theorem 5.3, so SLA invariants, which are common in
databases, might still make the encoding undecidable.  While this
concern can be addressed by considering machine (finite-bit)
arithmetic instead of SLA, we decided a better approach would be
revisit our encoding algorithm to see if it could be simplified.  We
were successful in this effort and were able to modify our encoding to
eliminate the existential altogether. Our modifications are based on
the observation that the semantics of bind, which were previously
written using two implications one of which involves an existential,
can be equivalently written using a single bi-implication that only
involves universal quantifiers. A proof of the equivalence between the
two encodings is given below; an automatic Z3-assisted proof, can be
found [here](https://rise4fun.com/Z3/G1a). As a consequence of this
change, the fragment of encoding is now unambiguously in EPR, and
hence Theorems 5.2 and 5.3 hold without the need for any modification
or qualification.  We believe this new development addresses the main
concerns of reviewer B, and trust that you will consider the revision
suitable for acceptance.

Sincerely,

Gowtham (for Kartik, Mahsa, and Suresh)

---------------------------------------

### 2017-10-22 popl-rebuttal.md

Shepard
----------
- The primary concern raised by reviewer B is that the fragment of FOL
  in which the translation algorithm (Fig. 10, Sec.5) encodes set
  expressions is not EPR, hence its decidability is doubtful. In
  particular, the concern was that an existential quantifier follows a
  universal quantifier in the logical encoding of bind (`>>=`), and
  this pattern is not admitted by EPR. In the rebuttal phase we
  addressed the concern by pointing out that the fragment, while not
  EPR as claimed, is nonetheless decidable since it is subsumed by the
  GKS fragment, which is known to be decidable. However, as the
  reviewer rightly points out, there are no known results on the
  combination of GKS and Simple Linear Arithmetic (SLA) to support
  theorem 5.3, so SLA invariants, which are common in databases, might
  make the encoding undecidable. While this concern can be addressed
  by considering machine (finite-bit) arithmetic instead of SLA, we
  decided to take a long, hard look at the encoding algorithm and see
  if it can be simplified. In this we were successful as we were able
  to modify our bind encoding to eliminate the existential altogether. Our
  modifications are based on the observation that the semantics of
  bind, which were previously written using two implications one of
  which involves an existential, can be equivalently written using a
  single bi-implication that only involves universal quantifiers. A
  formal proof to this effect is given below, while an automatic
  Z3-assisted proof can be found [here](https://rise4fun.com/Z3/G1a).
  As a consequence of this change, the fragment of encoding is back to
  being EPR, hence the Theorems 5.2 and 5.3 continue to hold. We have
  also tightened our treatment of free variables in the encoding
  algorithm, and added an example to Sec. 5.2 to demonstrate how
  exactly the encoding works.

- We added the following note to the end of the first paragraph of
  Sec. 3 to address reviewer C's comment about heap manipulating
  programs: "Our language ensures that all effectful actions are
  encapsulated within database commands, and that all shared state
  among processes are only manipulated via transactions and its
  supported operations.  In particular, we do not consider programs in
  which objects resident on e.g., the OCaml heap are concurrently
  manipulated by OCaml expressions as well as database actions."

### 2017-10-22 popl-rebuttal.md

- Dealing with heap state: It is indeed true that we have ignored heap
  memory to focus exclusively on database state in our
  formalization. Swamy et al. (PLDI'13), however, have proposed the
  Dijkstra monad to reason about heap memory monadically.  As part of
  future work, we intend to explore the composition of the DB and
  Dijkstra monads to enable reasoning over both heap and database
  state.

Shepherd
----------

Dear Peter -

Please find attached a revised near-final version of our POPL
submission.  Besides numerous minor changes throughout in response to
general themes raised in the reviews (e.g., the issue of of heap state
raised by reviewer (C) is addressed in Sec. 3 (last two sentences of
para. 1)), the most significant change from the original submission
deal with issues raised specifically by reviewer (B), which is
consistent with your guidance as well.

@@

Sincerely,

Gowtham (for Kartik, Mahsa, and Suresh)

---------------------------------------

Theorem:
  Let S be a set containing elements of type T, and let R be a binary
  relation over T. Let S1 be a T set defined using the following two
  implications (note: (x,y: T) is a shorthand for (x:T, y:T)):

           ∀(x,y: T). x∈S ∧ R(x,y) ⇒ y∈S1			(Call it H0)
           ∀(y: T). y∈S1 ⇒ ∃(x:T). x∈S ∧ R(x,y)		(H1)

  Likewise, let S2 be a T set defined via a single bi-implication as
  following:

          ∀(x,y: T). x∈S ∧ R(x,y) ⇔ y∈S2			(H2)

  Then, S1 = S2. That is:

          ∀(y: T). y∈S1 ⇔ y∈S2        (G)

Proof:
  We will show G by proving the following two implications (goals)
  separately:

        ∀(y: T). y∈S1 ⇒ y∈S2					(G0)
        ∀(y: T). y∈S2 ⇒ y∈S1					(G1)

  Let us first note that H2 is a conjunction of the following two
  hypothesis:

          ∀(x,y: T). x∈S ∧ R(x,y) ⇒ y∈S2			(H3)
          ∀(x,y: T). y∈S2 ⇒ x∈S ∧ R(x,y)			(H4)

  Proof of G0:
  Intros on the goal G0 introduces a new y0:T, gives us the following
  hypothesis:

          y0∈S1							(H5)

  And reduces the goal to proving that y0∈S2.  Instantiating H1 with
  (y:=y0), and then applying it on H5 gives:

          ∃(x:T). x∈S ∧ R(x,y0)				(H6)

  Destructing the existential in H6 introduces a new x0:T, and gives the
  following:

          x0∈S ∧ R(x0,y0)				(H7)

  Now, instantiating H3 with (x:=x0)(y:=y0), and applying it on H7 gives
  y0∈S2, which is what needs to be proven.

  Proof of G1:
  Intros on the goal G1 introduces a new y0:T, gives us the following
  hypothesis:

          y0∈S2					(H8)

  And reduces the goal to proving that y0∈S1. Instantiating H4 with
  (x:=y0)(y:=y0), and then applying it on H8 gives:

          y0∈S ∧ R(y0,y0)				(H9)

  Finally, instantiating H0 with (x:=y0)(y:=y0), and applying it on H9
  gives y0∈S1, which is what needs to be proven.
Qed.

### 2017-10-22 popl-rebuttal.md

- Dealing with heap state: It is indeed true that we have ignored heap
  memory to focus exclusively on database state in our
  formalization. Swamy et al. (PLDI'13), however, have proposed the
  Dijkstra monad to reason about heap memory monadically.  As part of
  future work, we intend to explore the composition of the DB and
  Dijkstra monads to enable reasoning over both heap and database
  state.

Shepherd
----------

Dear Peter -

Please find attached a revised near-final version of our POPL
submission.  Besides numerous minor changes throughout in response to
general themes raised in the reviews (e.g., the issue of heap state
raised by reviewer (C) is addressed in Sec. 3 (last two sentences of
para. 1)), the most significant changes from the original submission
deal with issues raised specifically by reviewer (B), which is
consistent with your guidance as well.

The primary concern raised by the reviewer is that the fragment of FOL
in which the translation algorithm (Fig. 10, Sec.5) encodes set
expressions is not EPR, and hence this raises doubts about its
decidability.  In particular, the concern was that an existential
quantifier follows a universal quantifier in the logical encoding of
bind (`>>=`), and this pattern is not admitted by EPR. In our
rebuttal, we addressed the concern by pointing out that the fragment,
while not EPR as claimed, is nonetheless decidable since it is
subsumed by GKS (Godel-Kalmar-Schutte), which is known to be
decidable. However, as the reviewer correctly points out, there are no
known results on the combination of GKS and Simple Linear Arithmetic
(SLA) to support theorem 5.3, so SLA invariants, which are common in
databases, might still make the encoding undecidable.  While this
concern can be addressed by considering machine (finite-bit)
arithmetic instead of SLA, we decided a better approach would be
revisit our encoding algorithm to see if it could be simplified.

We are pleased to report that we were successful in this effort and
were able to modify our encoding to eliminate the existential
altogether. Our modifications are based on the observation that the
semantics of bind, which were previously written using two
implications one of which involves an existential, can be equivalently
written using a single bi-implication that only involves universal
quantifiers. A proof of the equivalence between the two encodings is
given below; a mechanized Z3-assisted proof can be found at:
https://rise4fun.com/Z3/G1a. As a consequence of this change, the
fragment of encoding is now unambiguously in EPR, and hence Theorems
5.2 and 5.3 hold without the need for any modification or
qualification.  We believe this new development addresses the main
concerns of reviewer B, and consequently trust that you will consider
the revision suitable for acceptance.

### 2017-10-22 inference.tex

The state transformer rules, like the earlier RG rules, closely follow
the corresponding reduction rules in Fig.~\ref{fig:txnimp}, except
that their language of expression is $\SL$. For instance, while the
reduction rule for \C{UPDATE} declaratively specifies the set of
updated records, the state transformer rule uses $\SL$'s bind
operation to \emph{compute} the set. Other SQL rules do likewise. The
rules for \C{LET} binders, conditionals, and sequences compose the
effects inferred for their subcommands. Thus, the effect of a sequence
of commands $c_1;c_2$ is the union of effects $F_1$ and $F_2$ of $c_1$
and $c_2$, respectively, except that $F_2$ is applied to a state
($\stl$) to which $F_1$ has already been applied, reflecting their
order of reduction. The inference rule for \C{FOREACH} takes advantage
of the $\SL$'s bind operator to lift the effect inferred for the loop
body to the level of the loop. Since records added to $\stl$ in each
iteration of \C{FOREACH} are independent of the previous iteration,
sequential composition of the effects of different iterations is same
as their parallel composition. Since the loop body is executed once
per each $z\in x$, the effect of the the loop is the union of effects
($\F$) for all $z\in x$, all applied to the same state ($\stl$ and
$\stg$). That is, $\F_{loop}(\stl,\stg) = \bigcup_{z\in
  x}\F_{body}(\stl,\stg)$. From the definition of the set monad's bind
operator, $\F_{loop}(\stl,\stg) = x \bind (\lambda
z.~F_{body}(\stl,\stg))$, which mirrors the definition of the rule.
This development mirrors ideas found in type checkers for certain
dependent type systems~\cite{KJ14}.

>>>>

The state transformer rules, like the earlier RG rules, closely follow
the corresponding reduction rules in Fig.~\ref{fig:txnimp}, except
that their language of expression is $\SL$. For instance, while the
reduction rule for \C{UPDATE} declaratively specifies the set of
updated records, the state transformer rule uses $\SL$'s bind
operation to \emph{compute} the set. Other SQL rules do likewise. The
rules for \C{LET} binders, conditionals, and sequences compose the
effects inferred for their subcommands. Thus, the effect of a sequence
of commands $c_1;c_2$ is the union of effects $F_1$ and $F_2$ of $c_1$
and $c_2$, respectively, except that $F_2$ is applied to a state
($\stl$) to which $F_1$ has already been applied, reflecting their
order of reduction. The inference rule for \C{FOREACH} takes advantage
of the $\SL$'s bind operator to lift the effect inferred for the loop
body to the level of the loop. Since records added to $\stl$ in each
iteration of \C{FOREACH} are independent of the previous iteration (the
rule assumes local context independence as discussed in \S\ref{sec:opsem}),
the sequential composition of the effects of different iterations is the same
as their parallel composition. Since the loop body is executed once
per each $z\in x$, the effect of the the loop is the union of effects
($\F$) for all $z\in x$, all applied to the same state ($\stl$ and
$\stg$). That is, $\F_{loop}(\stl,\stg) = \bigcup_{z\in
  x}\F_{body}(\stl,\stg)$. From the definition of the set monad's bind
operator, $\F_{loop}(\stl,\stg) = x \bind (\lambda
z.~F_{body}(\stl,\stg))$, which mirrors the definition of the rule.
%This development mirrors ideas found in type checkers for certain
%dependent type systems~\cite{KJ14}.

### 2017-10-22 introduction.tex

Unfortunately, weak isolation admits behaviors that are difficult to
comprehend~\cite{berenson}. To quantify weak isolation anomalies,
Fekete \emph{et al.}~\cite{feketevldb09} devised and experimented with
a microbenchmark suite that executes transactions under a
weakly-isolated \emph{read committed} isolation level - the default
level for 8 of the 18 databases studied in~\cite{bailishotos}, and
found that 25 out of every 1000 rows in the database violate at least
one integrity constraint. Bailis \emph{et al.}~\cite{bailisferal} rely
on Rails' \emph{uniqueness validation} to maintain uniqueness of
records while serving Linkbench's~\cite{linkbench} insertion workload
(6400 records distributed over 1000 keys; 64 concurrent clients), and
report discovering more than 10 duplicate records.  Rails relies on
database transactions to validate uniqueness during insertions, which
is sensible if transactions are serializable, but incorrect under the
weak isolation level used in the experiments. The same study has found
that 13\% of all invariants among 67 open source Ruby-on-Rails
applications are at risk of being violated due to weak
isolation. Indeed, incidents of safety violations due to executing
applications in a weakly-isolated environment have been reported on
web services in production~\cite{starbucksbug, scimedbug}, including
in safety-critical applications such as bitcoin
exchanges~\cite{poloniexbug, bitcoinbug}. While enforcing
serializability for all transactions would be sufficient to avoid
these errors and anomalies, it would likely be an overly conservative
strategy; indeed, 75\% of the invariants studied in~\cite{bailisferal}
were shown to be preserved under some form of weak isolation.  When to
use weak isolation, and in what form, is therefore a prominent
question facing all database programmers.\footnote{This position has
been echoed by leading database researchers who lament the lack of a
better understanding of this problem; see e.g., {\tt
  http://www.bailis.org/blog/understanding-weak-isolation-is-a-serious-problem}.}

>>>>

Unfortunately, weak isolation admits behaviors that are difficult to
comprehend~\cite{berenson}. To quantify weak isolation anomalies,
Fekete \emph{et al.}~\cite{feketevldb09} devised and experimented with
a microbenchmark suite that executes transactions under a
weakly-isolated \emph{read committed} isolation level - the default
level for 8 of the 18 databases studied in~\cite{bailishotos}, and
found that 25 out of every 1000 rows in the database violate at least
one integrity constraint. Bailis \emph{et al.}~\cite{bailisferal} rely
on Rails' \emph{uniqueness validation} to maintain uniqueness of
records while serving Linkbench's~\cite{linkbench} insertion workload
(6400 records distributed over 1000 keys; 64 concurrent clients), and
report discovering more than 10 duplicate records.  Rails relies on
database transactions to validate uniqueness during insertions, which
is sensible if transactions are serializable, but incorrect under the
weak isolation level used in the experiments. The same study has found
that 13\% of all invariants among 67 open source Ruby-on-Rails
applications are at risk of being violated due to weak
isolation. Indeed, incidents of safety violations due to executing
applications in a weakly-isolated environment have been reported on
web services in production~\cite{starbucksbug, scimedbug}, including
in safety-critical applications such as bitcoin
exchanges~\cite{poloniexbug, bitcoinbug}. While enforcing
serializability for all transactions would be sufficient to avoid
these errors and anomalies, it would likely be an overly conservative
strategy; indeed, 75\% of the invariants studied in~\cite{bailisferal}
were shown to be preserved under some form of weak isolation.  When to
use weak isolation, and in what form, is therefore a prominent
question facing all database programmers.\footnote{This position has
been echoed by database researchers who lament the lack of a
better understanding of this problem; see e.g., {\tt
  http://www.bailis.org/blog/understanding-weak-isolation-is-a-serious-problem}.}

### 2017-10-22 opsem.tex
Typos and cosmetic changes.

We define a small-step operational semantics for this language in
terms of an abstract machine that executes a command, and updates
either a transaction-local ($\stl$), or global ($\stg$) database, both
of which are modeled as a set of records of a pre-defined type, i.e.,
they all belong to a single table.  The generalization to multiple
tables is straightforward, e.g., by having the machine manipulate a
set of sets, one for each table.  The semantics assumes that records
in $\stg$ can be uniquely identified via their $\idf$ field, and
enforces this property wherever necessary. Certain hidden fields are
treated specially by the operational semantics, and are hidden
from the surface language. These include a $\txnf$ field that tracks
the identifier of the transaction that last updated the record, and a
$\delf$ field that flags deleted records in $\stl$.  For a set $S$ of
records, we define $\dom(S)$ as the set of unique ids of all records
in $S$. Thus $|\dom(\stg)| = |\stg|$. During its execution, a
transaction may write to multiple records in $\stg$. Atomicity
dictates that such writes should not be visible in $\stg$ until the
transaction commits. We therefore associate each transaction with a
local database ($\stl$) that stores such uncommitted
records\footnote{While SQL's \C{UPDATE} admits writes at the
  granularity of record fields, most popular databases enforce
  record-level locking, allowing us to think of ``uncommitted writes''
  as ``uncommitted records''. }. Uncommitted records include deleted
records, whose $\delf$ field is set to \C{true}. When the transaction
commits, its local database is atomically \emph{flushed} to the global
database, committing these heretofore uncommitted records. The flush
operation ($\rhd$) is defined as follows:
\begin{smathpar}
\begin{array}{c}
\forall r.~ r \in (\stl\rhd\stg) ~\Leftrightarrow~ 
  (r.\idf \notin \dom(\stl) \conj r \in \stg)
\disj (r \in \stl \conj \neg r.\delf) 
\end{array}
\end{smathpar}
Let $\stg' = \stl\rhd\stg$. A record $r$ belongs to $\stg'$ iff it
belongs to $\stg$ and has not been updated in $\stl$, i.e., $r.\idf
\notin \dom(\stl)$, or it belongs to $\stl$, i.e., it is either a new
record, or an updated version of an old record, provided the update is
not a deletion ($\neg r.\delf$).  Besides the commit, flush also helps
a transaction read its own writes. Intuitively, the result of a read
operation inside a transaction must be computed on the database
resulting from flushing the current local state ($\stl$) to the global
state ($\stg$). The abstract machine of Fig.~\ref{fig:txnimp},
however, does not let a transaction read its own writes. This
simplifies the semantics, without losing any generality, since
substituting $\stl\rhd\stg$ for $\stg$ at select places in the
reduction rules effectively allows reads of uncommitted transaction
writes to be realized.

@@

The reduction of \C{FOREACH} starts by first converting it to its
run-time form to keep track of iterated items ($s_1$), as well as
yet-to-be-iterated items ($s_2$).  Iteration involves invoking its
function argument with $s_1$ and the current element $x$ (note:
$\uplus$ in $\{x\} \uplus s_2$ denotes a disjoint union). The
reduction ends when $s_2$ becomes empty. The reduction rules for
conditionals, \C{LET} binders, and sequences are standard, and
ommitted for brevity.

@@

\textbf{Local Context Independence} As mentioned previously, our
operational semantics does not let a transaction read its own writes.
It also does not let a transaction overwrite its own writes, due to
the premise $\dom(\stl)\cap\dom(s) = \emptyset$ on the
\rulelabel{E-Delete} and \rulelabel{E-Update} rules. We refer to this
restriction as \emph{local context independence}.  This restriction is
easy to relax in the operational semantics and the reasoning framework
presented in the next section; our inference procedure described in
\S\ref{sec:inference}, however, has a non-trivial dependence on
this assumption. Nonetheless, we have encountered few instances in
practice where enforcing local context independence turns out to be a
severe restriction. Indeed, all 17 transactions in the TPC-C and TPC-E
benchmarks satisfy this assumption.

>>>>

We define a small-step operational semantics for this language in
terms of an abstract machine that executes a command, and updates
either a transaction-local ($\stl$), or global ($\stg$) database, both
of which are modeled as a set of records of a pre-defined type, i.e.,
they all belong to a single table.  The generalization to multiple
tables is straightforward, e.g., by having the machine manipulate a
set of sets, one for each table.  The semantics assumes that records
in $\stg$ can be uniquely identified via their $\idf$ field, and
enforces this property wherever necessary. Certain hidden fields are
treated specially by the operational semantics, and are hidden
from the surface language. These include a $\txnf$ field that tracks
the identifier of the transaction that last updated the record, and a
$\delf$ field that flags deleted records in $\stl$.  For a set $S$ of
records, we define $\dom(S)$ as the set of unique ids of all records
in $S$. Thus $|\dom(\stg)| = |\stg|$. During its execution, a
transaction may write to multiple records in $\stg$. Atomicity
dictates that such writes should not be visible in $\stg$ until the
transaction commits. We therefore associate each transaction with a
local database ($\stl$) that stores such uncommitted
records\footnote{While SQL's \C{UPDATE} admits writes at the
  granularity of record fields, most popular databases enforce
  record-level locking, allowing us to think of ``uncommitted writes''
  as ``uncommitted records''. }. Uncommitted records include deleted
records, whose $\delf$ field is set to \C{true}. When the transaction
commits, its local database is atomically \emph{flushed} to the global
database, committing these heretofore uncommitted records. The flush
operation ($\rhd$) is defined as follows:
\begin{smathpar}
\begin{array}{c}
\forall r.~ r \in (\stl\rhd\stg) ~\Leftrightarrow~ 
  (r.\idf \notin \dom(\stl) \conj r \in \stg)
\disj (r \in \stl \conj \neg r.\delf) 
\end{array}
\end{smathpar}
Let $\stg' = \stl\rhd\stg$. A record $r$ belongs to $\stg'$ iff it
belongs to $\stg$ and has not been updated in $\stl$, i.e., $r.\idf
\notin \dom(\stl)$, or it belongs to $\stl$, i.e., it is either a new
record, or an updated version of an old record, provided the update is
not a deletion ($\neg r.\delf$).  Besides the commit, flush also helps
a transaction read its own writes. Intuitively, the result of a read
operation inside a transaction must be computed on the database
resulting from flushing the current local state ($\stl$) to the global
state ($\stg$). The abstract machine of Fig.~\ref{fig:txnimp},
however, does not let a transaction read its own writes. This
simplifies the semantics, without losing any generality, since
substituting $\stl\rhd\stg$ for $\stg$ at select places in the
reduction rules effectively allows reads of uncommitted transaction
writes to be realized, if so desired.

@@

The reduction of \C{FOREACH} starts by first converting it to its
run-time form to keep track of iterated items ($s_1$), as well as
yet-to-be-iterated items ($s_2$).  Iteration involves invoking its
function argument with $s_1$ and the current element $x$ (note:
$\uplus$ in $\{x\} \uplus s_2$ denotes a disjoint union). The
reduction ends when $s_2$ becomes empty. The reduction rules for
conditionals, \C{LET} binders, and sequences are standard, and
omitted for brevity.

@@

\textbf{Local Context Independence} As mentioned previously, our
operational semantics does not let a transaction read its own writes.
It also does not let a transaction overwrite its own writes, due to
the premise $\dom(\stl)\cap\dom(s) = \emptyset$ on the
\rulelabel{E-Delete} and \rulelabel{E-Update} rules. We refer to this
restriction as \emph{local context independence}.  This restriction is
easy to relax in the operational semantics and the reasoning framework
presented in the next section; our inference procedure described in
\S\ref{sec:inference}, however, has a non-trivial dependence on this
assumption.  Nonetheless, we have encountered few instances in
practice where enforcing local context independence turns out to be a
severe restriction. Indeed, all of the transactions we have considered
in our benchmarks (e.g., TPC-C) satisfy this assumption.
